Simple repression draft

### 2020-06-05 doc/appendix_01_nonbursty.tex
Not quite final pass on everything, Appdx S3 still needs more

\subsection{Derivation of chemical master equation}

@@

Since mRNA production can only take place in the transcriptionally active state
we will focus on this function for our derivation. The repressor bound state
will have an equivalent equation without terms involving the production of
mRNAs. We begin by listing the possible state transition that can happen for a
particular mRNA count with the promoter in the active state. For state changes
in a small time window $\Delta t$ that ``jump into'' state $m$ in the
transcriptionally active state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m-1$ to $m$.
    \item Degrade an mRNA, jumping from $m+1$ to $m$.
    \item Transition from the repressor bound state $R$ with $m$ mRNAs  to the
    active state $A$ with $m$ mRNAs.
\end{itemize}
Likewise, for state transitions that ``jump out'' of state $m$ in the
transcriptionally inactive state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m$ to $m+1$.
    \item Degrade an mRNA, jumping from $m$ to $m-1$.
    \item Transition from the active state $A$ with $m$ mRNAs to the repressor
    bound state $R$ with $m$ mRNAs.
\end{itemize}
The mRNA production does not depend on the current number of mRNAs, therefore
these state transitions occur with probability $r\Delta t$. The same is true for
the promoter state transitions, each occurs with probability $k_R^\pm \Delta t$.
As for the mRNA degradation events, these transitions depend on the current
number of mRNA molecules since the more molecules of mRNA there are, the more
will decay during a given time interval. Each molecule has a constant
probability $\gamma \Delta t$ of being degraded, so the total probability for an
mRNA degradation event to happen is computed by multiplying this probability by
the current number of mRNAs.

@@

All we have to do now are simple algebraic steps in order to simplify the 
equations. Let's focus on the transcriptionally active state $A$. First we will
send the term $p_A(m, t)$ to the right-hand side, and then we will divide both
sides of the equation by $\Delta t$. This results in
\begin{equation}
\begin{aligned}
\frac{p_A(m, t + \Delta t) - p_A(m, t)}{\Delta t} &=
r p_A(m-1, t) - r p_A(m, t)\\
&+ (m+1)\gamma p_A(m+1, t)
- m \gamma p_A(m, t)\\
&+k_R^- p_R(m, t)
-k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
Upon taking the limit when $\Delta t \rightarrow 0$ we can transform the 
left-hand side into a derivative, obtaining the chemical master equation
\begin{equation}
\begin{aligned}
\frac{d p_A(m, t)}{dt} &=
r p_A(m-1, t) - r p_A(m, t)\\
&+ (m+1)\gamma p_A(m+1, t)
- m \gamma p_A(m, t)\\
&+k_R^- p_R(m, t)
-k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
Doing equivalent manipulations for the repressor bound state gives an ODE of the
form
\begin{equation}
\begin{aligned}
\frac{d p_R(m, t)}{dt} &=
(m+1)\gamma p_R(m+1, t)
- m \gamma p_R(m, t)\\
&-k_R^- p_R(m, t)
+k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
In the next section we will write these coupled ODEs in a more compact form using
matrix notation.

@@

Along the way it will be convenient to define the following
\textit{conditional} moments as
\begin{equation}
\langle\vec{m}\rangle = \sum_{m=0}^\infty m \vec{p}(m),
\end{equation}
and
\begin{equation}
\langle\vec{m}^2\rangle = \sum_{m=0}^\infty m^2 \vec{p}(m).
\end{equation}
These objects are vectors of the same size as $\vect{p}(m)$, and each component
can be thought of as the expected value of the mRNA copy number, or copy number
squared, conditional on the promoter being in a certain state. For example, for
model one in Figure~\ref{fig1:means_cartoons} which has repressor bound and
unbound states labeled $R$ and $U$, $\langle\vec{m}^2\rangle$ would be
\begin{equation}
\langle\vec{m}^2\rangle
= \begin{pmatrix} \sum_{m=0}^\infty m^2 p_R(m)
                \\ \sum_{m=0}^\infty m^2 p_U(m) \end{pmatrix}.
\end{equation}
Analogously to $\langle\vec{m}\rangle$ and $\langle\vec{m}^2\rangle$,
it is convenient to define the vector
\begin{equation}
\langle\vec{m}^0\rangle = \sum_{m=0}^\infty \vec{p}(m),
\end{equation}
whose elements are simply the probabilities of finding the system in each of the
possible promoter states. To reduce notational clutter in what follows, sums
over $m$ will be assumed to run from 0 to $\infty$ unless otherwise specified,
i.e., $\sum_{m=0}^\infty$ will be denoted simply as $\sum_m$. It will also be
convenient to denote by $\vec{1}^\dagger$ a row vector of the same length as
$\vec{p}$ whose elements are all 1, such that, for instance, $\vec{1}^\dagger
\cdot \langle\vec{m}^0\rangle = 1$, $\vec{1}^\dagger \cdot \langle\vec{m}\rangle
= \langle{m}\rangle$, etc.

\subsubsection{Promoter state probabilities $\langle\vec{m}^0\rangle$}
\label{sec:m0_appdx}
To begin, we will find the promoter state probabilities
$\langle\vec{m}^0\rangle$ from Eq.~\ref{eq:generic_cme_ss} by summing over all
mRNA copy numbers $m$, producing
\begin{equation}
0 = \sum_m \left[
    \left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
                + \mathbf{R_A} \vec{p}(m-1) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1)
\right]
\end{equation}
Using the definitions of $\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$,
and noting the matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$
are all independent of $m$ and can be moved outside the sum, this simplifies to
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \sum_m \vec{p}(m-1)
    + \gamma \sum_m (m+1)\vec{p}(m+1).
\label{eq:generic_cme_deriv_020}
\end{equation}
The last two terms can be handled by reindexing the summations, transforming
them to match the definitions of $\langle\vec{m}^0\rangle$ and
$\langle\vec{m}\rangle$. For the first, noting $\vec{p}(-1)=0$ since negative
numbers of mRNA are nonsensical, we have
\begin{equation}
\sum_{m=0}^\infty \vec{p}(m-1)
= \sum_{m=-1}^\infty \vec{p}(m)
= \sum_{m=0}^\infty \vec{p}(m) = \langle\vec{m}^0\rangle.
\end{equation}
Similarly for the second,
\begin{equation}
\sum_{m=0}^\infty (m+1)\vec{p}(m+1)
= \sum_{m=1}^\infty m\vec{p}(m)
= \sum_{m=0}^\infty m\vec{p}(m) = \langle\vec{m}\rangle,
\end{equation}
which holds since in extending the lower limit from $m=1$ to $m=0$, the extra
term we added to the sum is zero. Substituting these back in
Eq.~\ref{eq:generic_cme_deriv_020}, we have
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \langle\vec{m}^0\rangle
    + \gamma \langle\vec{m}\rangle,
\end{equation}
or simply
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle.
\label{eq:generic_cme_vecm0}
\end{equation}
So given matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$ describing a
promoter, finding $\langle\vec{m}^0\rangle$ simply amounts to solving this set
of linear equations, subject to the normalization constraint $\vec{1}^\dagger
\cdot \langle\vec{m}^0\rangle = 1$. It will turn out to be the case that, if the
matrix $\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}$ is $n$ dimensional, it will
always have only $n-1$ linearly independent equations. Including the
normalization condition provides the $n$-th linearly independent equation,
ensuring a unique solution. So when using this equation to solve for
$\langle\vec{m}^0\rangle$, we may always drop one row of the matrix equation at
our convenience and supplement the system with the normalization condition
instead.

\subsubsection{First moments $\langle\vec{m}\rangle$ and $\langle{m}\rangle$}
By analogy to the above procedure for finding $\langle\vec{m}^0\rangle$, we may
find $\langle\vec{m}\rangle$ by first multiplying Eq.~\ref{eq:generic_cme_ss} by
$m$ and then summing over $m$. Carrying out this procedure we have
\begin{equation}
0 = \sum_m m \left[
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
            + \mathbf{R_A} \vec{p}(m-1) +
            \gamma (m+1) \mathbf{I} \vec{p}(m+1)
\right],
\end{equation}
and now identifying $\langle\vec{m}\rangle$ and $\langle\vec{m}^2\rangle$ gives
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}\rangle
    - \gamma \langle\vec{m}^2\rangle + \mathbf{R_A} \sum_m m\vec{p}(m-1)
    + \gamma \sum_m m(m+1)\vec{p}(m+1).
\label{eq:generic_cme_deriv_030}
\end{equation}
The summations in the last two terms can be reindexed just as we did for
$\langle\vec{m}^0\rangle$, freely adding or removing terms from the sum which
are zero. For the first,
\begin{equation}
\sum_{m=0}^\infty m\vec{p}(m-1)
= \sum_{m=-1}^\infty (m+1)\vec{p}(m)
= \sum_{m=0}^\infty (m+1)\vec{p}(m)
= \langle\vec{m}\rangle + \langle\vec{m}^0\rangle,
\end{equation}
and for the second,
\begin{equation}
\sum_{m=0}^\infty m(m+1)\vec{p}(m+1)
= \sum_{m=1}^\infty (m-1)m\vec{p}(m)
= \sum_{m=0}^\infty (m-1) m\vec{p}(m)
= \langle\vec{m}^2\rangle - \langle\vec{m}\rangle.
\end{equation}
Substituting back in Eq.~\ref{eq:generic_cme_deriv_030} then produces
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}\rangle
- \gamma \langle\vec{m}^2\rangle
+ \mathbf{R_A} (\langle\vec{m}\rangle + \langle\vec{m}^0\rangle)
+ \gamma (\langle\vec{m}^2\rangle - \langle\vec{m}\rangle),
\end{equation}
or after simplifying
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - \gamma) \langle\vec{m}\rangle
+ \mathbf{R_A} \langle\vec{m}^0\rangle.
\label{eq:generic_cme_deriv_040}
\end{equation}
So like $\langle\vec{m}^0\rangle$, $\langle\vec{m}\rangle$ is also found by
simply solving a set of linear equations after first solving for
$\langle\vec{m}^0\rangle$ from Eq.~\ref{eq:generic_cme_vecm0}.

@@

\subsubsection{Second moment $\langle{m}^2\rangle$ and Fano factor $\nu$}
Continuing the pattern of the zeroth and first moments, we now find
$\langle\vec{m}^2\rangle$ by multiplying Eq.~\ref{eq:generic_cme_ss} by $m^2$
and then summing over $m$, which explicitly is
\begin{equation}
0 = \sum_m m^2 \left[
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
            + \mathbf{R_A} \vec{p}(m-1) +
            \gamma (m+1) \mathbf{I} \vec{p}(m+1)
\right].
\end{equation}
Identifying the moments $\langle\vec{m}^2\rangle$ and $\langle\vec{m}^3\rangle$
in the first term simplifies this to
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^2\rangle
    - \gamma \langle\vec{m}^3\rangle + \mathbf{R_A} \sum_m m^2\vec{p}(m-1)
    + \gamma \sum_m m^2(m+1)\vec{p}(m+1).
\label{eq:generic_cme_deriv_050}
\end{equation}
Reindexing the sums of the last two terms proceeds just as it did for the zeroth
and first moments. Explicitly, we have
\begin{equation}
\sum_{m=0}^\infty m^2\vec{p}(m-1)
= \sum_{m=-1}^\infty (m+1)^2\vec{p}(m)
= \sum_{m=0}^\infty (m+1)^2\vec{p}(m)
= \langle\vec{m}^2\rangle + 2\langle\vec{m}\rangle + \langle\vec{m}^0\rangle,
\end{equation}
for the first sum and
\begin{equation}
\sum_{m=0}^\infty m^2(m+1)\vec{p}(m+1)
= \sum_{m=1}^\infty (m-1)^2m\vec{p}(m)
= \sum_{m=0}^\infty (m-1)^2 m\vec{p}(m)
= \langle\vec{m}^3\rangle - 2\langle\vec{m}^2\rangle + \langle\vec{m}\rangle
\end{equation}
for the second. Substituting the results of the sums back in
Eq.~\ref{eq:generic_cme_deriv_050} gives
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^2\rangle
- \gamma \langle\vec{m}^3\rangle
+ \mathbf{R_A}
    (\langle\vec{m}^2\rangle + 2\langle\vec{m}\rangle + \langle\vec{m}^0\rangle)
+ \gamma
    (\langle\vec{m}^3\rangle - 2\langle\vec{m}^2\rangle + \langle\vec{m}\rangle),
\end{equation}
and after grouping like powers of $m$ we have
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - 2\gamma) \langle\vec{m}^2\rangle
+ (2\mathbf{R_A} + \gamma) \langle\vec{m}\rangle
+ \mathbf{R_A} \langle\vec{m}^0\rangle.
\label{eq:generic_cme_deriv_060}
\end{equation}
As we found when computing $\langle{m}\rangle$ from $\langle\vec{m}\rangle$, we
can spare ourselves some algebra by multiplying
Eq.~\ref{eq:generic_cme_deriv_060} by $\vect{1}^\dagger$, which then reduces to
\begin{equation}
0 = - 2\gamma \langle{m}^2\rangle
+ \vec{1}^\dagger\cdot(2\mathbf{R_A} + \gamma) \langle\vec{m}\rangle
+ \vec{1}^\dagger\cdot\mathbf{R_A} \langle\vec{m}^0\rangle,
\end{equation}
and noting from Eq.~\ref{eq:generic_mean_m_appdx} that
$\vec{1}^\dagger\cdot\mathbf{R_A} \langle\vec{m}^0\rangle
= \gamma\langle{m}\rangle$, we have the tidy result
\begin{equation}
\langle{m}^2\rangle
= \langle{m}\rangle + \frac{1}{\gamma}
        \vec{1}^\dagger\cdot\mathbf{R_A} \langle\vec{m}\rangle.
\end{equation}

@@

First we need $\langle\vec{m}^0\rangle$. Label its components as $p_R$ and
$p_U$, the probabilities of finding the system in either promoter state, and
note that only $p_U$ survives multiplication by $\mathbf{R}$, since
\begin{equation}
\mathbf{R} \langle\vec{m}^0\rangle
= \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix}
    \begin{pmatrix} p_R \\ p_U \end{pmatrix}
= \begin{pmatrix} 0 \\ r p_U \end{pmatrix},
\label{eq:Rm0_model1_appdx}
\end{equation}
so we need not bother finding $p_R$. Then we have
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle
= \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix}
    \begin{pmatrix} p_R \\ p_U \end{pmatrix} = 0.
\end{equation}
As mentioned earlier in Section~\ref{sec:m0_appdx}, the two rows are linearly
dependent, so taking only the first row and using normalization to set $p_R =
1-p_U$ gives
\begin{equation}
-k_R^- (1-p_U) + k_R^+ p_U = 0,
\end{equation}
which is easily solved to find
\begin{equation}
p_U = \frac{k_R^-}{k_R^- + k_R^+}.
\end{equation}
Substituting this into Eq.~\ref{eq:Rm0_model1_appdx}, and the result of that
into Eq.~\ref{eq:mean_m_appdx_ref}, we have
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma} \frac{k_R^-}{k_R^- + k_R^+}
\end{equation}
as asserted in Eq.~\ref{eq:mean_m_model1} of the main text.

>>>>

\subsection{Derivation of chemical master equation}
\label{sec:cme_from_cartoon}

@@

Since mRNA production can only take place in the transcriptionally active state
we will focus on this function for our derivation. The repressor bound state
will have an equivalent equation without terms involving the production of
mRNAs. We begin by listing the possible state transitions that can occur for a
particular mRNA count with the promoter in the active state. For state changes
in a small time window $\Delta t$ that ``jump into'' state $m$ in the
transcriptionally active state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m-1$ to $m$.
    \item Degrade an mRNA, jumping from $m+1$ to $m$.
    \item Transition from the repressor bound state $R$ with $m$ mRNAs  to the
    active state $A$ with $m$ mRNAs.
\end{itemize}
Likewise, for state transitions that ``jump out'' of state $m$ in the
transcriptionally inactive state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m$ to $m+1$.
    \item Degrade an mRNA, jumping from $m$ to $m-1$.
    \item Transition from the active state $A$ with $m$ mRNAs to the repressor
    bound state $R$ with $m$ mRNAs.
\end{itemize}
The mRNA production does not depend on the current number of mRNAs, therefore
these state transitions occur with probability $r\Delta t$. The same is true for
the promoter state transitions; each occurs with probability $k_R^\pm \Delta t$.
As for the mRNA degradation events, these transitions depend on the current
number of mRNA molecules since the more molecules of mRNA there are, the more
will decay during a given time interval. Each molecule has a constant
probability $\gamma \Delta t$ of being degraded, so the total probability for an
mRNA degradation event to occur is computed by multiplying this probability by
the current number of mRNAs.

@@

All we have to do now are simple algebraic steps in order to simplify the 
equations. Let us focus on the transcriptionally active state $A$. First we will
send the term $p_A(m, t)$ to the right-hand side, and then we will divide both
sides of the equation by $\Delta t$. This results in
\begin{equation}
\begin{aligned}
\frac{p_A(m, t + \Delta t) - p_A(m, t)}{\Delta t} &=
r p_A(m-1, t) - r p_A(m, t)\\
&+ (m+1)\gamma p_A(m+1, t)
- m \gamma p_A(m, t)\\
&+k_R^- p_R(m, t)
-k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
Upon taking the limit when $\Delta t \rightarrow 0$ we can transform the 
left-hand side into a derivative, obtaining the chemical master equation
\begin{equation}
\begin{aligned}
\frac{d p_A(m, t)}{dt} &=
r p_A(m-1, t) - r p_A(m, t)\\
&+ (m+1)\gamma p_A(m+1, t)
- m \gamma p_A(m, t)\\
&+k_R^- p_R(m, t)
-k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
Doing equivalent manipulations for the repressor bound state gives an ODE of the
form
\begin{equation}
\begin{aligned}
\frac{d p_R(m, t)}{dt} &=
(m+1)\gamma p_R(m+1, t)
- m \gamma p_R(m, t)\\
&-k_R^- p_R(m, t)
+k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
In the next section we will write these coupled ODEs in a more compact form using
matrix notation.

@@

Along the way it will be convenient to define the following
\textit{conditional} moments as
\begin{equation}
\langle\vec{m}\rangle = \sum_{m=0}^\infty m \vec{p}(m),
\end{equation}
and
\begin{equation}
\langle\vec{m}^2\rangle = \sum_{m=0}^\infty m^2 \vec{p}(m).
\end{equation}
These objects are vectors of the same size as $\vect{p}(m)$, and each component
can be thought of as the expected value of the mRNA copy number, or copy number
squared, conditional on the promoter being in a certain state. For example, for
model one in Figure~\ref{fig1:means_cartoons} which has repressor bound and
unbound states labeled $R$ and $U$, $\langle\vec{m}^2\rangle$ would be
\begin{equation}
\langle\vec{m}^2\rangle
= \begin{pmatrix} \sum_{m=0}^\infty m^2 p_R(m)
                \\ \sum_{m=0}^\infty m^2 p_U(m) \end{pmatrix}.
\end{equation}
Analogously to $\langle\vec{m}\rangle$ and $\langle\vec{m}^2\rangle$,
it is convenient to define the vector
\begin{equation}
\langle\vec{m}^0\rangle = \sum_{m=0}^\infty \vec{p}(m),
\end{equation}
whose elements are simply the probabilities of finding the system in each of the
possible promoter states. It will be
convenient to denote by $\vec{1}^\dagger$ a row vector of the same length as
$\vec{p}$ whose elements are all 1, such that, for instance, $\vec{1}^\dagger
\cdot \langle\vec{m}^0\rangle = 1$, $\vec{1}^\dagger \cdot \langle\vec{m}\rangle
= \langle{m}\rangle$, etc.

\subsubsection{Promoter state probabilities $\langle\vec{m}^0\rangle$}
\label{sec:m0_appdx}
To begin, we will find the promoter state probabilities
$\langle\vec{m}^0\rangle$ from Eq.~\ref{eq:generic_cme_ss} by summing over all
mRNA copy numbers $m$, producing
\begin{equation}
0 = \sum_{m=0}^\infty \left[
    \left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
                + \mathbf{R_A} \vec{p}(m-1) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1)
\right]
\end{equation}
Using the definitions of $\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$,
and noting the matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$
are all independent of $m$ and can be moved outside the sum, this simplifies to
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \sum_{m=0}^\infty \vec{p}(m-1)
    + \gamma \sum_{m=0}^\infty (m+1)\vec{p}(m+1).
\label{eq:generic_cme_deriv_020}
\end{equation}
The last two terms can be handled by reindexing the summations, transforming
them to match the definitions of $\langle\vec{m}^0\rangle$ and
$\langle\vec{m}\rangle$. For the first, noting $\vec{p}(-1)=0$ since negative
numbers of mRNA are nonsensical, we have
\begin{equation}
\sum_{m=0}^\infty \vec{p}(m-1)
= \sum_{m=-1}^\infty \vec{p}(m)
= \sum_{m=0}^\infty \vec{p}(m) = \langle\vec{m}^0\rangle.
\end{equation}
Similarly for the second,
\begin{equation}
\sum_{m=0}^\infty (m+1)\vec{p}(m+1)
= \sum_{m=1}^\infty m\vec{p}(m)
= \sum_{m=0}^\infty m\vec{p}(m) = \langle\vec{m}\rangle,
\end{equation}
which holds since in extending the lower limit from $m=1$ to $m=0$, the extra
term we added to the sum is zero. Substituting these back in
Eq.~\ref{eq:generic_cme_deriv_020}, we have
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \langle\vec{m}^0\rangle
    + \gamma \langle\vec{m}\rangle,
\end{equation}
or simply
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle.
\label{eq:generic_cme_vecm0}
\end{equation}
So given matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$ describing a
promoter, finding $\langle\vec{m}^0\rangle$ simply amounts to solving this set
of linear equations, subject to the normalization constraint $\vec{1}^\dagger
\cdot \langle\vec{m}^0\rangle = 1$. It will turn out to be the case that, if the
matrix $\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}$ is $n$ dimensional, it will
always have only $n-1$ linearly independent equations. Including the
normalization condition provides the $n$-th linearly independent equation,
ensuring a unique solution. So when using this equation to solve for
$\langle\vec{m}^0\rangle$, we may always drop one row of the matrix equation at
our convenience and supplement the system with the normalization condition
instead.
The reader may find it illuminating to skip ahead and see
Eq.~\ref{eq:generic_cme_vecm0} in use with concrete examples, e.g.,
Eq.~\ref{eq:model1_m0_giver_appdx} and what follows it, before
continuing on through the general formulas for moments.

\subsubsection{First moments $\langle\vec{m}\rangle$ and $\langle{m}\rangle$}
By analogy to the above procedure for finding $\langle\vec{m}^0\rangle$, we may
find $\langle\vec{m}\rangle$ by first multiplying Eq.~\ref{eq:generic_cme_ss} by
$m$ and then summing over $m$. Carrying out this procedure we have
\begin{equation}
0 = \sum_{m=0}^\infty m \left[
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
            + \mathbf{R_A} \vec{p}(m-1) +
            \gamma (m+1) \mathbf{I} \vec{p}(m+1)
\right],
\end{equation}
and now identifying $\langle\vec{m}\rangle$ and $\langle\vec{m}^2\rangle$ gives
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}\rangle
    - \gamma \langle\vec{m}^2\rangle + \mathbf{R_A} \sum_{m=0}^\infty m\vec{p}(m-1)
    + \gamma \sum_{m=0}^\infty m(m+1)\vec{p}(m+1).
\label{eq:generic_cme_deriv_030}
\end{equation}
The summations in the last two terms can be reindexed just as we did for
$\langle\vec{m}^0\rangle$, freely adding or removing terms from the sum which
are zero. For the first term we find
\begin{equation}
\sum_{m=0}^\infty m\vec{p}(m-1)
= \sum_{m=-1}^\infty (m+1)\vec{p}(m)
= \sum_{m=0}^\infty (m+1)\vec{p}(m)
= \langle\vec{m}\rangle + \langle\vec{m}^0\rangle,
\end{equation}
and similarly for the second,
\begin{equation}
\sum_{m=0}^\infty m(m+1)\vec{p}(m+1)
= \sum_{m=1}^\infty (m-1)m\vec{p}(m)
= \sum_{m=0}^\infty (m-1) m\vec{p}(m)
= \langle\vec{m}^2\rangle - \langle\vec{m}\rangle.
\end{equation}
Substituting back in Eq.~\ref{eq:generic_cme_deriv_030} then produces
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}\rangle
- \gamma \langle\vec{m}^2\rangle
+ \mathbf{R_A} (\langle\vec{m}\rangle + \langle\vec{m}^0\rangle)
+ \gamma (\langle\vec{m}^2\rangle - \langle\vec{m}\rangle),
\end{equation}
or after simplifying
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - \gamma) \langle\vec{m}\rangle
+ \mathbf{R_A} \langle\vec{m}^0\rangle.
\label{eq:generic_cme_deriv_040}
\end{equation}
So like $\langle\vec{m}^0\rangle$, $\langle\vec{m}\rangle$ is also found by
simply solving a set of linear equations after first solving for
$\langle\vec{m}^0\rangle$ from Eq.~\ref{eq:generic_cme_vecm0}.

@@

\subsubsection{Second moment $\langle{m}^2\rangle$ and Fano factor $\nu$}
Continuing the pattern of the zeroth and first moments, we now find
$\langle\vec{m}^2\rangle$ by multiplying Eq.~\ref{eq:generic_cme_ss} by $m^2$
and then summing over $m$, which explicitly is
\begin{equation}
0 = \sum_{m=0}^\infty m^2 \left[
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
            + \mathbf{R_A} \vec{p}(m-1) +
            \gamma (m+1) \mathbf{I} \vec{p}(m+1)
\right].
\end{equation}
Identifying the moments $\langle\vec{m}^2\rangle$ and $\langle\vec{m}^3\rangle$
in the first term simplifies this to
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^2\rangle
    - \gamma \langle\vec{m}^3\rangle + \mathbf{R_A} \sum_{m=0}^\infty m^2\vec{p}(m-1)
    + \gamma \sum_{m=0}^\infty m^2(m+1)\vec{p}(m+1).
\label{eq:generic_cme_deriv_050}
\end{equation}
Reindexing the sums of the last two terms proceeds just as it did for the zeroth
and first moments. Explicitly, we have
\begin{equation}
\sum_{m=0}^\infty m^2\vec{p}(m-1)
= \sum_{m=-1}^\infty (m+1)^2\vec{p}(m)
= \sum_{m=0}^\infty (m+1)^2\vec{p}(m)
= \langle\vec{m}^2\rangle + 2\langle\vec{m}\rangle + \langle\vec{m}^0\rangle,
\end{equation}
for the first sum and
\begin{equation}
\sum_{m=0}^\infty m^2(m+1)\vec{p}(m+1)
= \sum_{m=1}^\infty (m-1)^2m\vec{p}(m)
= \sum_{m=0}^\infty (m-1)^2 m\vec{p}(m)
= \langle\vec{m}^3\rangle - 2\langle\vec{m}^2\rangle + \langle\vec{m}\rangle
\end{equation}
for the second. Substituting the results of the sums back in
Eq.~\ref{eq:generic_cme_deriv_050} gives
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^2\rangle
- \gamma \langle\vec{m}^3\rangle
+ \mathbf{R_A}
    (\langle\vec{m}^2\rangle + 2\langle\vec{m}\rangle + \langle\vec{m}^0\rangle)
+ \gamma
    (\langle\vec{m}^3\rangle - 2\langle\vec{m}^2\rangle + \langle\vec{m}\rangle),
\end{equation}
and after grouping like powers of $m$ we have
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - 2\gamma) \langle\vec{m}^2\rangle
+ (2\mathbf{R_A} + \gamma) \langle\vec{m}\rangle
+ \mathbf{R_A} \langle\vec{m}^0\rangle.
\label{eq:generic_cme_deriv_060}
\end{equation}
As we found when computing $\langle{m}\rangle$ from $\langle\vec{m}\rangle$, we
can spare ourselves some algebra by multiplying
Eq.~\ref{eq:generic_cme_deriv_060} by $\vect{1}^\dagger$, which then reduces to
\begin{equation}
0 = - 2\gamma \langle{m}^2\rangle
+ \vec{1}^\dagger\cdot(2\mathbf{R_A} + \gamma) \langle\vec{m}\rangle
+ \vec{1}^\dagger\cdot\mathbf{R_A} \langle\vec{m}^0\rangle,
\end{equation}
and noting from Eq.~\ref{eq:generic_mean_m_appdx} that
$\vec{1}^\dagger\cdot\mathbf{R_A} \langle\vec{m}^0\rangle
= \gamma\langle{m}\rangle$, we have the tidy result
\begin{equation}
\langle{m}^2\rangle
= \langle{m}\rangle + \frac{1}{\gamma}
        \vec{1}^\dagger\cdot\mathbf{R_A} \langle\vec{m}\rangle.
\end{equation}

@@

First we need $\langle\vec{m}^0\rangle$. Label its components as $p_R$ and
$p_U$, the probabilities of finding the system in either promoter state, and
note that only $p_U$ survives multiplication by $\mathbf{R}$, since
\begin{equation}
\mathbf{R} \langle\vec{m}^0\rangle
= \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix}
    \begin{pmatrix} p_R \\ p_U \end{pmatrix}
= \begin{pmatrix} 0 \\ r p_U \end{pmatrix},
\label{eq:Rm0_model1_appdx}
\end{equation}
so we need not bother finding $p_R$. Then we have
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle
= \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix}
    \begin{pmatrix} p_R \\ p_U \end{pmatrix} = 0.
\label{eq:model1_m0_giver_appdx}
\end{equation}
As mentioned earlier in Section~\ref{sec:m0_appdx}, the two rows are linearly
dependent, so taking only the first row and using normalization to set $p_R =
1-p_U$ gives
\begin{equation}
-k_R^- (1-p_U) + k_R^+ p_U = 0,
\end{equation}
which is easily solved to find
\begin{equation}
p_U = \frac{k_R^-}{k_R^- + k_R^+}.
\end{equation}
Substituting this into Eq.~\ref{eq:Rm0_model1_appdx}, and the result of that
into Eq.~\ref{eq:mean_m_appdx_ref}, we have
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma} \frac{k_R^-}{k_R^- + k_R^+}
\end{equation}
as asserted in Eq.~\ref{eq:mean_m_model1} of the main text.
